\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation

### 2020-01-13 2.ResearchReport/Presentation/Presentation.Rmd
Rename folders to make structure more clear, adjust readme file in 'main' repository.

## Multiple imputation
- Missing data is ubiquitous [@alli02]. 
- *Ad hoc* solutions may yield invalid inferences [@buur18].
- @rubin87 proposed the framework of MI.

@@

## Algorithmic convergence
<center>
<img src="figures/non-conv.png" alt="HTML5 Icon" width = 77%>
<br>

<!-- Iteration number -->
</center>

@@

## Results | Simulation diagnostics versus convergence diagnostics
```{r echo=FALSE}
dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE)
```

@@

## Equations $\widehat{R}$
\[\begin{align}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text{where} \quad \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \\
&\quad \text{ and } \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\

\widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\quad \text{where} \quad \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{align}\]

>>>>

## Multiple imputation
- General statement:

\[\text{Missing data is ubiquitous.}\] 
<br>

- *Ad hoc* solutions may yield invalid inferences [@buur18].
<br>

- @rubin87 proposed the framework of MI.

@@

## Algorithmic convergence
<center>
<img src="figures/non-conv.png" alt="HTML5 Icon" width = 77%>
<br>
<!-- Iteration number -->
</center>

## Equations $\widehat{R}$
\[\begin{align}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text{where} \quad \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \\
&\quad \text{ and } \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\
\widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\quad \text{where} \quad \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{align}\]

@@

## Results | Simulation diagnostics versus convergence diagnostics
```{r echo=FALSE}
dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 10), rownames = FALSE)
```

@@

### 2020-01-13 3.Thesis/article.Rnw
Rename folders to make structure more clear, adjust readme file in 'main' repository.

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}

@@

Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1. The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).

@@

\section{Simulation study}

\subsection{Simulation: Introduction}

@@

R hat not working on MI data? "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. Aim to replicate Lacerda et al. that R hat will not be smaller than 1.1 before iteration nr. 50.

@@

R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution). With MI procedures, initial values are chosen such that ... or maybe even estimated from the observed data. This means that at most one initial value is necessary, but probably none at al. I should look this up.

@@

Explain diagnostics here (see below).

We want to evaluate whether the imputation procedure has converged. We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance). The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}. That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.

Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise. There is no tradeoff between them. Also look at distributional char., and plausibility of imp. values, see \cite{vinknd}.

@@

Set-up: simulate data, impute missing data points with varying number of iterations. Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations. Convergence is evaluated with $\hat{R}$, or potential scale reduction factor. Proposed by Gelman and Rubin (1992), but since developed into ...
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)

In this study, 1000 MCMC simulations are performed.

\begin{Code}
# pseudocode of similation 
simulate data with missingness
for (number of iterations from 1 to 100)
  impute the missingness
  compute convergence diagnostic
  perform analysis
  pool results
  compute simulation diagnostics
save diagnostics
\end{Code}

Simulation code is available in online appendix XYZ on Github.

@@

Rhat is not perfect for MI data. The value shoudl theoretically not be smaller than one, yet it happened several times in this study. How could Rhat smaller than 1 occur? The number of simulations is smaller than in 'regular' MCMC processes. Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr. This downwards bias is in the opposite direction than expected.

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
% \title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
% \Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
% \Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Shorttitle{Convergence Diagnostics for MICE}

@@

Subsequently, I will use \proglang{R} Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1. The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).

@@

\section{Simulation study}% A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations}

\subsection{Introduction}

@@

Missing data is ubiquitous. MICE can solve stuff. [Copy Thesis Proposal Introduction.]

Validity of inference MCMC algorithms threatned by non-convergence. Non-converegnce has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains. Diagnose non-convergence: $\widehat{R}$, autocorrelation, MC error, Geweke, the other one. However, existing methodology may not work on MI data. Focus on $\widehat{R}$.

@@

Why is R hat not applicable on MI data? R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution). With MI procedures, initial values are chosen such that ... or maybe even estimated from the observed data. This means that at most one initial value is necessary, but probably none at al. I should look this up. And, "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. Empirical finding suggests otherwise: that R hat will not be smaller than 1.1 before iteration nr. 50. Therefore, the aim is to replicate empirical finding Lacerda et al.

@@

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether GR statistic Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. We can apply Rhat to the mean (or to the first two moments) of the variables of interest.  The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}. That is, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.

@@

In this study, 1000 MCMC simulations are performed. The simulation set-up consists of several steps, see pseudocode below. Simulation code is available in online appendix XYZ on Github.

\begin{Code}
# pseudocode of similation 
simulate data with missingness
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    impute the missingness
    compute convergence diagnostic
    perform analysis
    pool results
    compute simulation diagnostics
save convergence and simulation diagnostics
\end{Code}

The simulated data is a finite population of $N=1000$. The variables are dependent variable $Y$, independent variable $X$, and covariates $Z_1$ and $Z_2$. The quantity of scientific interest is the estimated regression coefficient of $X$ on $Y$ in linear regression model $Y \sim X+Z_1+Z_2$. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$. Outcome variable $Y$ is deduced from the predictor variables as follows:

\begin{align*}
\begin{pmatrix}X\\
Z_{1}\\
Z_{2}\\
\epsilon
\end{pmatrix} &\sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,&
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 10
\end{pmatrix}
\end{bmatrix}\\[2\jot]
Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
\end{align*}

After data generation, the complete data is 'amputed'. That is, the \pkg{mice} function \fct{ampute} is used to impose an MCAR missingness mechanism upon the data. The probability to be missing is the same for all cells, namely XYZ\%. Table 1 shows a summary of the generated complete data, and the amputed data. The amputed data is the starting point of each of the 1000 simulations.

Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains. The number of iterations is varied over simulations ('maxit' argument between 1 and 100). For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)

@@

Results over 1000 simulations are presented in Table XYZ. Rhat is not perfect for MI data. The value shoudl theoretically not be smaller than one, yet it happened several times in this study. How could Rhat smaller than 1 occur? The number of simulations is smaller than in 'regular' MCMC processes. Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction facotr. This downwards bias is in the opposite direction than expected.

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Tue Nov 12 17:43:11 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \hline
It. & Bias & Emp. SE & CI width & Cov. rate & Rhat & Autocorr. \\ 
  \hline
  1 & -0.135 & 0.115 & 0.637 & 0.930 &  &   \\ 
  2 & -0.088 & 0.125 & 0.691 & 0.990 & 1.499 & -0.500 \\ 
  3 & -0.090 & 0.112 & 0.623 & 0.950 & 1.200 & -0.658 \\ 
  4 & -0.093 & 0.116 & 0.645 & 0.980 & 1.146 & -0.738 \\ 
  5 & -0.087 & 0.108 & 0.599 & 0.980 & 1.098 & -0.711 \\ 
  6 & -0.092 & 0.116 & 0.644 & 0.950 & 1.090 & -0.674 \\ 
  7 & -0.095 & 0.122 & 0.675 & 1.000 & 1.063 & -0.588 \\ 
  8 & -0.099 & 0.113 & 0.626 & 0.970 & 1.058 & -0.523 \\ 
  9 & -0.095 & 0.115 & 0.639 & 1.000 & 1.044 & -0.519 \\ 
  10 & -0.090 & 0.111 & 0.618 & 0.990 & 1.048 & -0.414 \\ 
  15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
  20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
  25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
  30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
  40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
  50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
   \hline
\end{tabular}
\end{table}

And figure

% \includegraphics[height=80mm,keepaspectratio]{res3.pdf}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{res3.pdf}
  }      
  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
    \label{resultssim}
\end{figure}

### 2020-01-13 3.Thesis/article.tex
Rename folders to make structure more clear, adjust readme file in 'main' repository.

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
% \title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
% \Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
% \Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Shorttitle{Convergence Diagnostics for MICE}

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
